---
title: "randomisation"
author: "Brad Duthie"
date: "21/09/2020"
output: ioslides_presentation
runtime: shiny
---

<style type="text/css">
body p {
  color: #000000;
}
ul {
  color: black;
}
ol {
  color: black;
}
a {
    color: darkblue;
    text-decoration: none
}
.fifty {
   font-size: 50%;
   }
.seventy {
   font-size: 70%;
   }
slides > slide.backdrop {
  background: none !important;
  background-color: white !important;
}
slides > slide.title-slide hgroup h1 {
  color: black;
}
slides > slide.title-slide p {
  color: black;
}
slides > slide.light > hgroup h2 {
  color: black;
}
slides > slide.light > hgroup h1 {
  color: black;
}
ul {
  color: black;
}
ol {
  color: black;
}
p {
  color: black
}
h1, h2, h3, h4, h5, h6, h7, h8, p {
    color: black;
}
.column-left{
  float: left;
  width: 67%;
  text-align: left;
}
.column-right{
  float: right;
  width: 33%;
  text-align: right;
}
</style>


<!----
<style type="text/css">
body p {
  color: #000000;
}
</style>
--->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## {.columns-2}

{.column}

```{r, echo=FALSE}
inputPanel(
  actionButton("go", "Randomise the data"),
  actionButton("stop", "Reset")
)
```


```{r, echo = FALSE, fig.height = 5}
renderPlot({
    hist(rnorm(100));
})
```

<br>


```{r, echo=FALSE}
renderTable({
    matrix(data = rnorm(20), ncol = 2);
})
```




##

This is regular text spanning the whole page. But here comes a two-column section.

:::::: {.columns}
::: {.column width="48%" data-latex="{0.48\textwidth}"}
This text is in the left column.
:::
::: {.column width="4%" data-latex="{0.04\textwidth}"}
\ 
<!-- an empty Div (with a white space), serving as
a column separator -->
:::
:::::: {.column width="48%" data-latex="{0.48\textwidth}"}
This text is in the right column.
:::
::::::
\newline

And back to a regular single-column environment.
