---
title: "Introduction to randomisation, bootstrap, and Mote Carlo methods in R"
author: "Brad Duthie"
date: "23 January 2019"
output: 
    html_document:
        code_folding: show
bibliography: randomisation_notes.bib
---

Contents
================================================================================

********************************************************************************

**Note: Please think of this document as a living document, which you are free to improve (like a [wiki](https://en.wikipedia.org/wiki/Wiki)) with minor edits, new sections that others might find useful, or [additional resources](#whatelse) that you find that you think others might also find useful. After reading through this, you will have been introduced to some key statistical techniques in randomisation and Monte Carlo methods that can be coded in R.**

********************************************************************************

- [Introduction: Randomisation and Monte Carlo](#intro)
- [Random number generation in R: key functions](#rand_r) 
    - [Random normal: `rnorm`](#rnorm)
    - [Random uniform: `runif`](#runif)
    - [Random binomial: `runif`](#rbinom)
    - [Random poisson: `rpois`](#rpois)
    - [Random sampled values: `sample`](#sample)
- [Randomisation for hypothesis testing](#rand_h)
- [Bootstrapping confidence intervals](#rand_b)
- [Monte Carlo method for spatial data](#rand_m)

********************************************************************************

<a name="intro">Introduction: Randomisation and Monte Carlo</a>
================================================================================


The objective of these notes is to help the reader get started with using randomisation, bootstrap, and Monte Carlo methods in R.

<a name="rnorm">**Sampling from a normal distribution**</a>

The `rnorm` function returns some number (`n`) of (pseudo)randomly generated numbers given a set mean ($\mu$; `mean`) and standard deviation ($\sigma$; `sd`), such that $X\sim\mathcal N(\mu,\sigma^2)$. The default is to draw from a standard normal distribution (i.e., $\mu = 0$ and $\sigma = 1$).

```{r}
rand_norms_10 <- rnorm(n = 10, mean = 0, sd = 1);
```

The above code stores a vector of ten numbers, shown below.

```{r, echo = FALSE}
print(rand_norms_10);
```

We can verify that a standard normal distribution is generated by plotting a histogram of a very large number of values created using `rnorm`.

```{r}
rand_norms_10000 <- rnorm(n = 10000, mean = 0, sd = 1);
hist(rand_norms_10000, xlab = "Random value (X)", col = "grey",
     main = "", cex.lab = 1.5, cex.axis = 1.5);
```

<a name="runif">**Sampling from a uniform distribution**</a>

Like the `rnorm` function, the `runif` function returns some number (`n`) of random numbers, but from a uniform distribution with between a range of $a$ (`min`) and $b$ (`max`) such that $X\sim\mathcal U(a,b)$, where $-\infty < a < b < \infty$. The default is to draw from a standard uniform distribution (i.e., $a = 0$ and $b = 1$).

```{r}
rand_unifs_10 <- runif(n = 10, min = 0, max = 1);
```

The above code stores a vector of ten numbers, shown below.

```{r, echo = FALSE}
print(rand_unifs_10);
```

As with the randomly generated normally distributed numbers, we can verify that a standard uniform distribution is generated by plotting a histogram of a very large number of values created using `runif`.

```{r}
rand_unifs_10000 <- runif(n = 10000, min = 0, max = 1);
hist(rand_unifs_10000, xlab = "Random value (X)", col = "grey",
     main = "", cex.lab = 1.5, cex.axis = 1.5);
```

<a name="rbinom">**Sampling from a binomial distribution**</a>


<a name="rpois">**Sampling from a poisson distribution**</a>

<a name="sample">**Sampling values from a vector**</a>


<a name="rand_r">Random number generation in R: key functions</a>
================================================================================




<a name="rand_h">Randomisation for hypothesis testing</a>
================================================================================



<a name="rand_b">Bootstrapping confidence intervals</a>
================================================================================


<a name="rand_b">Monte Carlo method for spatial data</a>
================================================================================


<a name="whatelse">Additional resources</a>
================================================================================




References
================================================================================

